---
title: "R-tutorial block course"
author: "Angel Rain"
date: "3/10/2022"
output:
  html_document:
    toc: yes
    toc_float: no
  pdf_document:
    toc: yes
---
# Introduction
## Basic commands
```{r }
# Set a directory
# setwd() #<- Put the address here
# get current directory
getwd()
# Screen element in the workspace
ls()
# Remove element from the workspace
## Some
rm()
## All
rm(list=ls())
```
## Basic object
Data can be stored in several ways depending of the complexity or the nature of the dataset.

```{r }
#Vector
x=1:5 

## Generate random numbers
set.seed(1)
y=rnorm(10,5,1)

#Matrix
M=matrix(1:50,ncol=5,nrow=10)
M
Mt=t(M)
Mt
#Data frames
DF=data.frame(Experiment="Test1", # Experiment name as string 
           Treatments=factor(rep(c("Substrate1","Substrate2"),each=15)), #Treatment description here 2 levels
           Values=c(rnorm(15,4,1),rnorm(15,8,1))) # Theoretical response growth (mg L-1 h-1)

## See the first rows
head(DF)
## For the last rows
tail(DF)
#Overview of the different columns
summary(DF)

#List
List.01=list(x,y,M,Mt,DF)
## Extract element from list
List.01[[1]]
List.01[[5]]
List.01[[5]]$Treatments
```

# Data analysis
Data analysis can be divided according the numbers of samples (ppt)

## One sample analysis
One-sample t-test/ is used to compare the mean of one sample to a known standard (or theoretical/hypothetical) average
```{r,fig.align = 'center'}
y=rnorm(35,mean=5)
# Visualize data
## Histogram
par(mfrow=c(1,2))
hist(y,12, ylim=c(0,10))
abline(v=5,lty=2)
boxplot(y)
t.test(y) # by default test if mean is equal to 0 (P<<0.05)
t.test(y, mu=5) # Here P>0.5, then mean is equal to 5

```
## Two samples analysis
```{r, fig.align = 'center'}
## Visualize data
#install.packages("ggplot2")
#install.packages("cowplot")

library(ggplot2)
library(cowplot)

# Simple boxplots with different graphical options
p1=ggplot(DF,aes(x=Treatments, y=Values))+geom_boxplot()
# Simple boxplots with data points
p2=ggplot(DF,aes(x=Treatments, y=Values))+geom_boxplot(outlier.size = -1)+geom_jitter(width = 0.2)
# Simple boxplots colored
p3=ggplot(DF,aes(x=Treatments, y=Values, fill=Treatments))+geom_boxplot(outlier.size = -1)+ geom_jitter(width = 0.2)+theme(legend.position = "none")
# Simple boxplots with colored points
p4=ggplot(DF,aes(x=Treatments, y=Values, colour=Treatments))+geom_boxplot(outlier.size = -1)+ geom_jitter(width = 0.2)+theme(legend.position = "none")
plot_grid(p1,p2,p3,p4,labels = "AUTO",align = "v", axis = "l")

# t-test two samples
t.test(DF$Values[DF$Treatments=="Substrate1"], DF$Values[DF$Treatments=="Substrate2"],var.equal = T) #Default setup: Only diff

t.test(DF$Values[DF$Treatments=="Substrate1"], DF$Values[DF$Treatments=="Substrate2"],var.equal = T,alternative ="less")
#Alternative hypothesis sample_a<sample_b or differences <0

t.test(DF$Values[DF$Treatments=="Substrate1"], DF$Values[DF$Treatments=="Substrate2"],var.equal = T, alternative ="greater")
#Alternative hypothesis sample_a>sample_b or differences >0


```

## More than two samples
### ANOVA one-way
One way anovas considered continuous variables in discrete categories. For example, growth rates in different temperatures, respiration with different substrate as resources

```{r fig.align='center', fig.dim=c(5, 4), message=TRUE, warning=FALSE}
#Generate testing data frames
set.seed(100)
DF=data.frame(Experiment=as.character("Test1"), # Experiment name
           Treatments=factor(rep(c("Substrate1","Substrate2","Substrate3"),each=15)), #Treatment description here 2 levels
           Values=c(rnorm(15,4,1),rnorm(15,8,1),rnorm(15,8.5,1))) # Theoretical response growth (mg L-1 h-1)
# Visualizing using ggplot2 data set
ggplot(DF,aes(x=Treatments, y=Values, colour=Treatments))+ geom_boxplot(outlier.size = -1)+ geom_jitter(width = 0.2)+ theme(legend.position = "none")
```

### Statistical analysis
Loading packages for analysis

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(rstatix)
library(knitr)
library(kableExtra)
library(ggpubr)
```

### Assumptions and ANOVA
```{r message=FALSE}
#Normality of dataset
tapply(DF$Value,DF$Treatment,shapiro.test)

#Homogeneity of variance for each comparison (Treatment)
bartlett.test(Values~Treatments,data=DF)

#ANOVA one-way
anova1=DF%>%anova_test(Values ~ Treatments)
anova1
res.tukey=DF%>%tukey_hsd(Values ~ Treatments)
res.tukey <- res.tukey %>% add_xy_position(x = "Treatments")

```

### Visualization: box plots with p-values

```{r fig.align='center', fig.dim=c(5, 5),}
ggboxplot(DF, x = "Treatments", y = "Values",color="Treatments",add = "jitter") +
  stat_pvalue_manual(res.tukey, hide.ns = TRUE) +
  labs(subtitle = get_test_label(anova1, detailed = TRUE))
```

## ANOVA two-ways

```{r, fig.align = 'center' }
#Generate testing data frames
DF=data.frame(Experiment=as.character("Experiment01"), # Experiment name
           Treatments=factor(rep(c("Substrate1","Substrate2"),each=20)), #Treatment description here 2 levels
           Sites=factor(rep(c("River","Lake"),each=10)),
           Values=c(rnorm(10,15,1.1),rnorm(10,11.5,1.5),
                    rnorm(10,16,1),rnorm(10,11,1.2))) # Theoretical response growth (mg L-1 h-1)

# Visualizing data set
bxp <-ggboxplot(DF, x= "Sites", y = "Values",add = "jitter",
  color = "Treatments", palette = "jco")
bxp
```

## Assumptions and ANOVA
```{r message=FALSE}
#Normality of dataset
tapply(DF$Value,interaction(DF$Treatment,DF$Sites),shapiro.test)

#Homogeneity of variance for each comparison (Treatment)
bartlett.test (Values~interaction(Treatments,Sites),data=DF)

#ANOVA one-way
anova2=DF%>%anova_test(Values ~ Treatments*Sites)
anova2
```
From the results above, What we can say about the main effects? Where are the differences among groups?

```{r}
DF$index=paste0(DF$Treatments,DF$Sites)
res.tukey=DF%>%
  group_by(Sites)%>%
tukey_hsd(Values ~ Treatments)
#Add position for plots

```
# Correlations
```{r, fig.align = 'center' }
DF2=data.frame(int.explorer=seq(25,95,length.out=26)+rnorm(26,1,2),
               USA.murders=seq(10000,20000,length.out=26)+rnorm(26,1000,200),
               year=2020:1995)
#Time series
ggplot(DF2)+geom_bar(aes(year,USA.murders),stat="identity",fill="blue")+
    geom_line(aes(x=year, y=200*int.explorer),stat="identity",color="red",size=2)+
    labs(title= "Murders in USA vs use of Internet Explorer", x="Year",y="Murders per year")+
    scale_y_continuous(sec.axis=sec_axis(~.*0.02,name="Percentage Use Internet Explorer"))

# Correlation between variables
ggplot(data=DF2, aes(int.explorer,USA.murders))+labs(x="Internet explorer",y="USA murders")+
  geom_point()+geom_smooth(method="lm",se=F)+geom_text(aes(label=year))

#Results correlations by Species
cor.test(~int.explorer+USA.murders,data=DF2,method = "pearson")
```
# Multivariate Statistic
In the nature we usually observed many populations living together in a limited space with limited resources. For many years researcher have studied what are the rules governing, community composition changes

```{r message=FALSE, warning=FALSE}
rm(list=ls())
library(reshape2)
library(ggplot2)
library(gridExtra)
library(vegan)
library(RColorBrewer)
library(phyloseq)

# Colorblind palette
cbp1 <- c("#999999","#FFDB6D", "#E69F00", "#56B4E9", "#009E73",
         "#F0E442", "#0072B2", "#D55E00", "#CC79A7","#293352")

# Load datasets
asv <- read.table('../data/dada.counts.ASV.1.txt', sep="\t" ,header = TRUE,row.names = 1) #Load raw count table
tax <- read.table('../data/dada.Taxonomy.ASV.tab', sep="\t" ,header = TRUE)

#Rarefaction of samples
raremax <- min(rowSums(asv))
Srare <- rrarefy(asv, raremax)
asv=t(Srare)
asv=prop.table(asv,2)
asv<- asv*100 #multiply by 100 to get % data
asv=data.frame(asv)
ASV=rownames(asv)
asv=cbind(ASV,asv)


# Combine relative count-table with taxonomic information
df1<-(data.frame(tax, asv))
df1=df1[,c("Kingdom",  "Phylum" ,  "Class" ,   "Order"  ,  "Family"  ,"Genus" ,"ASV", "SOLA1","SOLA2", "SOLA3","Sea1","Sea2","Sea3","SM1","SM2", "SM3","SS1","SS2","SS3")]
```


```{r}
#aggregate counts by order
df2<-aggregate(cbind(Sea1,Sea2,Sea3,
                     SM1,SM2,SM3,
                     SOLA1,SOLA2,SOLA3,
                     SS1,SS2,SS3) ~ Order,
               data = df1, sum, na.rm = TRUE)

order <- df2[,1] #vector with orders
rownames(df2) <- df2[,1] #row names
df2<-df2[,-1] #remove column with orders and keep only abundance data
colSums(df2) #test the sum of columns to see if values close to 1 are reached

```
```{r}
agg=df2
#change format of agg to samples by column
agg<-as.data.frame(agg)
rownames(agg) <- order #add order information
agg$Sum.agg<-rowSums(agg) # add column with counts across treatments for all orders

#test how many of the orders (ranked by abundance) are meaningful to not include to many orders but reach values possibly close to 1
colSums(agg[with(agg, order(-Sum.agg)),][1:10,]) 

#select the most abundant orders (top 10)
agg10<-agg[with(agg, order(-Sum.agg)),][1:10,1:12]
agg10$order <-rownames(agg10)
```

```{r, fig.align = 'center',fig.dim=c(6,5)}
#convert to long format
agg10.long<-melt(agg10, id.vars='order', variable.name = "treat")

positions <- c("SW-DOM","SW-DOM","SW-DOM",
               'S-DOM',"S-DOM","S-DOM",
               "F-SEA","F-SEA","F-SEA",
               "Source","Source","Source")

getPalette = colorRampPalette(brewer.pal(10, "Paired"))

plot0<-ggplot(agg10.long, aes(x = treat, y = value, fill = order)) +
  geom_bar(stat = "identity")+scale_fill_manual(values = cbp1)+
  scale_x_discrete(limits=c("SM1","SM2","SM3","SS1","SS2","SS3", "Sea1","Sea2","Sea3","SOLA1","SOLA2","SOLA3"),
                   labels =positions)+labs(y="Relative abundance", x=NULL)+
  theme_bw()+  theme(axis.text.y =element_text(face="bold",size=10,angle=0,hjust = 1)) +coord_flip()

plot0

```

## Principal coordinate analysis (PCOA)
Multidimensional scaling have been used to represent in a 2 or 3D space the multivariate changes in community composition. Imagine every species in a population is a single variable, then multidimensional scaling allow to reduce dimensions to the most important ones. This reduction is based in the weight of the variance explained by the relative abundance of the species.

```{r, fig.align = 'center' ,fig.dim=c(10,4)}
########################## PCoA
ps.rel = otu_table(df1[,8:19], taxa_are_rows = TRUE)
#ggplot
#all samples
ord.pcoa<- ordinate(ps.rel, method="PCoA", distance="bray")
ord.nmds<- ordinate(ps.rel, method="NMDS", distance="bray",autotransform=T)

df.ord.pcoa=data.frame(pco1=ord.pcoa$vectors[,1],pco2=ord.pcoa$vectors[,2],
                            DOM=c("Source","Source","Source",
                                  "SW","SW","SW",
                                  "DOM2","DOM2","DOM2",
                                  "DOM1","DOM1","DOM1"))
df.ord.nmds<-data.frame(pco1=ord.nmds$points[,1],pco2=ord.nmds$points[,2])
df.ord.nmds$DOM=df.ord.pcoa$DOM
# Ordering factors
df.ord.pcoa$DOM=ordered(df.ord.pcoa$DOM,levels=c("DOM1","DOM2","SW", "Source"))
df.ord.nmds$DOM=ordered(df.ord.pcoa$DOM,levels=c("DOM1","DOM2","SW", "Source"))

p2=ggplot(df.ord.pcoa,aes(x=pco1, y=pco2,colour=DOM))+
  geom_point(size=3)+
    labs(title="Principal cordinate analysis (PCoA)",x=paste0("PCOA1 [",round(ord.pcoa$values[2][1,]*100,0),"%]") ,
       y=paste0("PCOA2 [",round(ord.pcoa$values[2][2,]*100,0),"%]"))+
  theme_bw()

p3=ggplot(df.ord.nmds,aes(x=pco1, y=pco2,colour=DOM))+
  geom_point(size=3)+labs(title="Non metric multidimensional scaling (NMDS)",x="Dim1",y="Dim2")+
  theme_bw()

plot_grid(p2,p3)
```

